---
title: "Mapping distribution and diversity using high and low def countries/regions"
author: "Thomas L.P. Couvreur"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
      keep_md: true
      number_sections: true
      theme: readable
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=TRUE, warning=FALSE, message=FALSE)
```

> Here we shall learn how to:  
- import and plot different type of maps, either at country or regional levels  
- import distribution data and plot them onto the maps  

# Initial things to do

## Set working directory

This is your own working directory, on your PC, update path below

```
setwd("C:/Users/couvreur/Documents/R/mada_annon_R/")
```

## Load needed libraries

**optional**: Install them first if not done already  

```
install.packages(c("ConR", "devtools", "mapproj", "maps", "rnaturalearthdata",
                   "rgbif", "raster", "rnaturalearth", "tidyverse", "maptools", "ggmap"))
```
Load the libraries

```
library(ggmap)
library(ggplot2)
library(sf)
library(dplyr)
library(ggplot2)
library(scico)
library(rnaturalearth)
library(purrr)
library(smoothr)
library(rworldxtra)
library(tidyr)
```

```{r include = FALSE}
library(ggmap)
library(ggplot2)
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(purrr)
library(smoothr)
library(rworldxtra)
library(tidyr)
library(raster)
```


# Plot country or regional maps

Select what country you want here:

```{r}
country = "Madagascar"
```

Or select the region you want, using coordinates:

```{r}
xlat = c(40, 60)
ylong = c(-26, -12)
```

We can plot maps at two levels of precision in terms of countries: Low and high.  
Maps are plotted using `ggplot2`, see after.  

## Low precision maps

1) Import a lower definition map from `rnaturalearth` package

```{r}
worldMap <- ne_countries(scale = "medium", type = "countries", returnclass = "sf")
```

2) Filter the country you want, creates a *NCpoly_low* object

```{r}
CNpoly_low <- worldMap %>% filter(sovereignt == country)
```

3) Simple plot of the map using *plot()* function
```{r}
plot(CNpoly_low$geometry)
```

## High presicion maps

1) Import a higher definition map from `rworldxtra` package

This is part of the package. We need to transform it to a `sp` format after using *st_as_sp* function. This procudes a *CNhigh* object

```{r}
data(countriesHigh)
CNhigh = st_as_sf(countriesHigh)
```

2) Filter the country you want, creates a *NCpoly* object

```{r}
CNpoly <- CNhigh %>% filter(ADMIN == country)
```

## Plot *country* map with `ggplot2`

Plot the high definition country filtered above.

Set limits around country, creates a *limsCN* object

```{r}
limsCN <- st_buffer(CNpoly, dist = 0.7) %>% st_bbox
```

Plot map with different options that can be changed. 

```{r}
divpolPlot <-
  ggplot() +
  geom_sf(data = CNpoly) +
  coord_sf(
    xlim = c(limsCN["xmin"], limsCN["xmax"]),
    ylim = c(limsCN["ymin"], limsCN["ymax"])
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )
divpolPlot
```

## Plot *regional* map with `ggplot2`

Here we shall indicate a region using coordinates, rather than filtering for a country.  
We use the *CNhigh* object.  
Here coordinates for Madagascar. CAREFULL: *xlat* and *ylong* are defined above.

```{r}
divpolPlot <-
  ggplot() +
  geom_sf(data = CNhigh) +
  coord_sf(
    xlim = xlat,
    ylim = ylong
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )
divpolPlot
```

## Add some climatic data to the plots

Here we shall download data from WorldClim and plot it with the map.

### Download or import climatic data from *WorldClim*

If already downloaded and strored on your PC locally:

```
 climate=getData('worldclim', var='bio',res=2.5, download=FALSE)
```

Otheriwe (this step can take some time (ca. 123 Mb of data)):

```{r}
climate=getData('worldclim', var='bio',res=2.5)
```

### Crop to region of interest

```{r}
climate <- crop(climate,extent(40, 60,-26, -12))
```

### Raster the layer you want, here *bio12*

```{r}
raster <- climate$bio12
```

### Clean na values
```{r}
rasdf <- as.data.frame(raster,xy=TRUE)%>%drop_na()
```

Now we have a rasterized layer of climate data (*bio12*) we can add to ggplot, see below.

# Plot species distribution data on map

## Import original raw occurence dataset

Update paths and file name to your own

```{r}
df <- read.csv("data/occs_cleaned_medrged_WAG_TAN_MO_P_03092021.csv", header = TRUE, sep = ";", stringsAsFactors = FALSE)
```

Our raw database contains **`r length(df$RecordedBy)`** unique records representing **`r length(unique(df$genus))`** genera and **`r length(unique(df$specific_epithet))`** species.

## Plot distribution data on *regional* map **without** climatic data

We have now added a *geom_point* argument to the ggplot script:  

`geom_point(data=df, aes(x=ddlong, y=ddlat), bg = rgb(red = 1, green = 0, blue = 0, alpha = 0.5), col = "black", pch = 21, cex=2) +`

Color and shape of points can be changed in the script.
In our file, the columns *ddlat* and *ddlong* indicate the coordinates.  

```{r}
divpolPlot <-
ggplot() +
  geom_sf(data = CNhigh) +
  geom_point(data=df, aes(x=ddlong, y=ddlat), bg = rgb(red = 1, green = 0, blue = 0, alpha = 0.5), col = "black", pch = 21, cex=2) +
  coord_sf(
    xlim = xlat,
    ylim = ylong
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )
divpolPlot + ggtitle("Distribution of Annonaceae species in Madagascar") + theme(plot.title = element_text(hjust = 0.5)) + xlab("Longitude") + ylab("Latitude")
```

## Plot distribution data on *regional* map **with** climatic data

This ggplot script is different than the one above as it adds the climate data.  

```{r}
ggplot()+
  geom_raster(aes(x=x,y=y,fill=bio12),data=rasdf)+
  geom_sf(fill='transparent',data=CNpoly)+
  scale_fill_viridis_c('mm/yr',direction = -1)+
  coord_sf(expand=c(0,0))+
  geom_point(data=df, aes(x=ddlong, y=ddlat), bg = rgb(red = 1, green = 0, blue = 0, alpha = 0.5), col = "black", pch = 21, cex=1) +
  coord_sf(
    xlim = xlat,
    ylim = ylong
  ) +
  labs(x='Longitude',y='Latitude',
       title="Annonaceae distribution in Madagascar's climate map",
       subtitle='Annual precipitation',
       caption='Source: WorldClim, 2020')+
  cowplot::theme_cowplot()+
  theme(panel.grid.major = element_line(color = gray(.5),
                                        linetype = 'dashed',
                                        size = 0.5),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill=NA,color = 'black'),
        panel.ontop = TRUE)
```


